Using best practices for forecasting biodiversity changes in breeding birds

### 2017-08-12 manuscript/draft.Rmd
table, other edits

@@

In this paper, we attempt to forecast the species richness of breeding birds at
over 1,000 of locations throughout North America while following the best
practices in Box 1 for ecological forecasting. To do this, we combine 30 years
of time-series data on bird distributions from annual surveys with monthly
time-series of climate data and satellite-based remote-sensing. This 30 year
time-scale has only recently become possible for large-scale time-series based
forecasting and allows us to model and assess changes a decade or more into the
future in the presence of shifts in environmental conditions on par with
predicted climate change. We compare traditional distribution modeling based
approaches to spatial models of species richness, time-series methods, and
simple baselines (Figure 1). All forecasts model uncertainty and observation
error, are evaluated across different time lags using hindcasting, and are
publicly archived to allow future assessment. We discuss the implications of
these practices for our understanding of, and confidence in, the resulting
forecasts, and how we can continue to build on these approaches improve
ecological forecasting in the future.

@@

Forecasting is an emerging imperative in ecology; as such, the field needs to
develop and follow best practices for conducting and evaluating ecological
forecasts. We have laid out ideas for a number of these practices (Box 1) and
attempted to implement them in a single study that builds and evaluates
forecasts of biodiversity in the form of species richness. The results of this
effort are both promising and humbling. On an absolute basis, when comparing
forecasts across sites, many different approaches to forecasting produce
reasonable forecasts. If a site is predicted to have a high number of species in
the future, relative to other sites, it generally does. However, the forecasts
perform relatively poorly at predicting variation in richness over time. If a
site is predicted to have a higher richness next year than it did last year,
there is little confidence that this will be the case. As a result, baseline
models, like the long-term average, along with the other site-level time-series
models, consistently provided the best forecasts for future biodiversity.

The most commonly used method to forecast future biodiversity, SDMs
[@urban2015], provided the worst overall forecasts of all methods evaluated. In
general, SDM-based forecasts are rarely evaluated, particularly in terms of
repeated observations at the same site. Since site-level richness is relatively
stable, model evaluations that show high accuracy across spatial gradients may
not indicate much about a model's ability to predict changes over time. This
issue has also recently been highlighted as an issue for using SDMs to predict
the future locations of individual species [@rapacciuolo2012]. This result is
particularly sobering because this approach forms the main foundation for
estimates of the predicted loss of biodiversity to climate change [@thomas2004;
@urban2015]. The poor performance of SDM-based forecasts highlights the crucial
importance of comparing multiple modeling approaches when conducting ecological
forecasts, and in particular the value of comparing results to simple baselines
to avoid over-interpreting the information present in these forecasts (Box 1).
Disciplines that have more mature forecasting cultures often do this by
reporting "forecast skill", i.e., the improvement in the forecast relative to a
simple baseline [@jolliffe2003]. We recommend following Ye et al. [@ye2015] and
adopting this approach in future ecological forecasting research. It also
highlights the need for using hindcasting and time-series data for assessing the
effectiveness of forecasting methods, since this allows comparisons to baselines
and assessment of predicted dynamics within sites or regions.

>>>>

```{r, echo=FALSE, message=FALSE}
library(tidyverse, quietly = TRUE)
library(kableExtra, quietly = TRUE)
```

@@

\begin{singlespace}

```{r table, echo=FALSE}

table_caption = "A summary of the six models evaluated here. The time series
models were trained site-by-site, without access to any environmental
information. The environmental models were trained at a continental scale,
without information regarding which transects occurred at which site or during
which year. Most of the models were trained to predict richness directly.
However, the random forest SDMs were each trained to predict the occurrence of
individual species at each site, and these species-level predictions were added
up to produce richness estimates. The mistnet JSDM was trained to predict the
full species composition at each site, and the number of species in its
predictions was used as an estimate of richness."

table_data = read_csv("../figures/table.csv",col_types = "ccccc")

table_data %>% 
  knitr::kable(
    format = "latex", 
    escape = FALSE, 
    booktabs = TRUE,
    caption = table_caption
  ) %>% 
  add_header_above(c(" " = 2, "Predictors" = 3)) %>% 
  group_rows("Time series models", 1, 3) %>% 
  group_rows("Environmental models", 4, 6) %>% 
  kable_styling(position = "center")
```

\end{singlespace}

In this paper, we attempt to forecast the species richness of breeding birds at
over 1,000 of locations throughout North America while following the best
practices in Box 1 for ecological forecasting. To do this, we combine 30 years
of time-series data on bird distributions from annual surveys with monthly
time-series of climate data and satellite-based remote-sensing. This 30 year
time-scale has only recently become possible for large-scale time-series based
forecasting and allows us to model and assess changes a decade or more into the
future in the presence of shifts in environmental conditions on par with
predicted climate change. We compare traditional distribution modeling based
approaches to spatial models of species richness, time-series methods, and
simple baselines (Table 1). All forecasts model uncertainty and observation
error, are evaluated across different time lags using hindcasting, and are
publicly archived to allow future assessment. We discuss the implications of
these practices for our understanding of, and confidence in, the resulting
forecasts, and how we can continue to build on these approaches improve
ecological forecasting in the future.

@@

Forecasting is an emerging imperative in ecology; as such, the field needs to
develop and follow best practices for conducting and evaluating ecological
forecasts. We have laid out ideas for a number of these practices (Box 1) and
attempted to implement them in a single study that builds and evaluates
forecasts of biodiversity in the form of species richness. The results of this
effort are both promising and humbling. When comparing forecasts across sites,
many different approaches to forecasting produce reasonable forecasts. If a site
is predicted to have a high number of species in the future, relative to other
sites, it generally does. However (after controlling for observer differences),
none of the methods we evaluated could reliably tell us whether site-level
richness would increase or decrease over time (Figure 6), which is generally the
stated purpose of these forecasts. As a result, our baseline models, which did
not attempt to anticipate changes in richness over time, generally provided
the best forecasts for future biodiversity.

These results raise a number of important issues. First, we should be skeptical
of predictions about future biodiversity from models that have not been
validated in terms of repeated observations at the same site. Since site-level
richness is relatively stable, model evaluations that show high accuracy across
spatial gradients may not indicate much about a model's ability to predict
changes over time. This issue has also recently been highlighted as an issue for
using SDMs to predict the future locations of individual species
[@rapacciuolo2012; @oedekoven2017]. This result is particularly sobering because
SDMs form the main foundation for estimates of the predicted loss of
biodiversity to climate change [@thomas2004; @urban2015].

Second, our results highlight the importance of comparing multiple modeling
approaches when conducting ecological forecasts, and in particular, the value of
comparing results to simple baselines to avoid over-interpreting the information
present in these forecasts (Box 1). Disciplines that have more mature
forecasting cultures often do this by reporting "forecast skill", i.e., the
improvement in the forecast relative to a simple baseline [@jolliffe2003]. We
recommend following Ye et al. [@ye2015] and adopting this approach in future
ecological forecasting research. It also highlights the need for using
hindcasting and time-series data for assessing the effectiveness of forecasting
methods, since this allows comparisons to baselines and assessment of predicted
dynamics within sites or regions.
